Cartogram Mapping and its Application to Cancer Data Visualization

### 2019-12-08 Response/Response.Rmd

Bowel Cancer Australia Atlas:
...shows the average Standardised Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia  . It is published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)*.

**Reviewer A: **
**Major**
*Comment 1: *
The Abstract should be more focused- perhaps give an example of how cartogram mapping has been applied to data visualisation. What are the benefits of this approach and suggestions for the future.

*Reply 1:* 
We agree with the reviewer that alternative approaches have been fantastic when used in political contexts. Choropleth maps have often been used.

*Changes in the text:* 
Cancer atlases communicate cancer statistics over geographic domains. These domains are subdivided by administrative areas such as countries, states or suburbs. When communicating human-related statistics in Australia, the geographic map base highlights the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they are important to consider if they are more densely populated. Alternative map displays can reveal patterns that are not obvious when the same data is shown on a choropleth map. Alternative displays have proven to be successful on online news sites, especially for the communication of election data, where votes are aggregated at the state level.

@@

*Comment 2: *

2. The key message seems to get lost in the Abstract. The Conclusion is vague.

*Reply 2:* 
See *Reply 1* for changes to the abstract. Additionally, the conclusion has also been edited to be more direct in concluding the alternatives presented.

### 2019-12-08 Response/Response.tex

Bowel Cancer Australia Atlas:
\ldots{}shows the average Standardised Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia . It is published by \emph{\href{https://www.bowelcanceraustralia.org/}{Bowel Cancer Australia}}.

\textbf{Reviewer A: }
\textbf{Major}
\emph{Comment 1: }
The Abstract should be more focused- perhaps give an example of how cartogram mapping has been applied to data visualisation. What are the benefits of this approach and suggestions for the future.

\emph{Reply 1:}
We agree with the reviewer that alternative approaches have been fantastic when used in political contexts. Choropleth maps have often been used.

\emph{Changes in the text:}
Cancer atlases communicate cancer statistics over geographic domains. These domains are subdivided by administrative areas such as countries, states or suburbs. When communicating human-related statistics in Australia, the geographic map base highlights the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they are important to consider if they are more densely populated. Alternative map displays can reveal patterns that are not obvious when the same data is shown on a choropleth map. Alternative displays have proven to be successful on online news sites, especially for the communication of election data, where votes are aggregated at the state level.

@@

\emph{Comment 2: }

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  The key message seems to get lost in the Abstract. The Conclusion is vague.
\end{enumerate}

\emph{Reply 2:}
See \emph{Reply 1} for changes to the abstract. Additionally, the conclusion has also been edited to be more direct in concluding the alternatives presented.

### 2019-12-08 review_article.Rmd
fix referencing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(cowplot)
library(png)
library(grid)
```
<!--

Please note that the submission deadline is September 30, 2019.

Author notes:
6000 words maximum.
focused  application  of  literature  to address  a  relevant  clinical  issue:
Mapping cancer outcomes
Cartogram applications should point to examples of cancer outcomes, with the main cartogram section discussing sources considered in grey literature review. Examples of political areas may be considered if cancer data is aggregated at these levels.

- Future steps should consider applications not currently utilized for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualizing those spatial patterns, or the uncertainty in those patterns
- a more applied-focused paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

\section*{Abstract}
<!-- 119 words -->
<!--
Choropleth maps are widely used in cancer atlases to communicate cancer statistics over geographic domains. The type of statistic displayed varies immensely, and effective choices can assist in the understanding of impact of cancer on a population. Atlases have moved from print media to web format, and the availability of graphical user interfaces, and user interaction and animation, greatly enhance the ability to communicate more information and positively engage readers. More use of alternatives to choropleth maps, such as cartograms, could be employed in online atlases, for better representation of spatial distributions, and show statistics on small areas. This is especially true for the geography of Australia, where the population is concentrated is small areas along the coastlines.  
-->

Cancer atlases communicate cancer statistics over geographic domains. These domains are often subdivided by administrative areas such as countries, states or suburbs.
When communicating human-related statistics in Australia, the geographic map base may highlight the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they may be important to consider if they are more densely populated. To

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the count data needs to be merged with population data to appropriately calibrate it to incidence, for example, the rate per 100000 people.  This type of data is collected on a routine basis for public health purposes and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of the disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

@@

# Map displays for disease data {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map [@EI]. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals in communities and organizing communities into representable units [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
@CPISACA discuss the use of choropleths to visualize cancer data, and @DMAHP gives an overview of the development of these maps for displaying disease data. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The data was extracted from the official federal statistics by @USInteractive on cancer incidence and deaths, produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. There is also a spatial outlier -- Utah has a noticeably lower rate than its neighbors. Also, Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

@@

# Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection of statistics across regions and the administrative control within regions lends itself to choropleth visualization. 
Atlases date to Haviland's maps in 1875, with UK cancer atlases appearing in 1961 [@HEDP], and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

@@

Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

@@

## Additional considerations
<!-- 479 words -->
Maps cannot tell the entire story. Supplementary graphs and plots are often included to add more depth and information. Additional materials such as tables, graphs, and text explanations, support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, standard deviation. When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses', and the display of the supplementary statistics helps in this regard.

@@

# Alternatives to choropleths {#ch:alternatives}

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that white lies may be employed to create useful displays and map creators can draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a valueother than the actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, interpreting spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, Dorling [@ACTUC] considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
<!-- @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. DI SAYS I think that focuses on bayesian analysis not on cartograms-->

@@

A contiguous cartogram maintains connectivity of the map regions while areas are altered according to a statistic. This transformation often occurs at the expense of the shape of areas [@CBATCC, @NAC, @TAAM].
From a computer graphics perspective, @ACA explain the application of 'map deformation' to account for the value assigned to each area, they provide three methods for creating value-by-area cartograms.
Examples include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].

@@

He acknowledged the sophistication of contiguous cartograms but critiqued their 'very complex shapes,' he answers this with his implementation of maps created using 'the simplest of all shapes'. 
Circular cartograms use the same simple shapes for every region represented and resize the shapes according to the statistic represented or the population for a base map. This familiar shape may be more effective for understanding the spatial distribution than contiguous cartograms, as the 'nonsense' shapes shown in contiguous cartograms have 'no meaning' after distortions are applied @NISCC.
To produce a compelling map, a gravity model is applied to avoid overlaps and keep spatial relationships with neighboring areas over many iterations. This implementation can work for up to 'one hundred thousand' areas.

@@

# Comparison and critique {#ch:compare}
<!-- 408 words -->

@@

Some mix of tiling, faceting or even micromaps, which allow some spatial continuity while also zooming into small areas, are good solutions for difficult geographies. Table \ref{tab:methods} summarizes the key criteria for evaluating maps and alternative displays. Suggestions and comments to help map creators best communicate their health data and spatial analyses are provided in @CPISACA and @SAMGIS.

@@

Monmonier [@HTLWM] says that interactivity can be used to allow users to explore the map for more information and provides flexibility for the display. The user can toggle between different variables, map views or even multiple realizations of future scenarios [@goodchild1994introduction]. This provides additional mechanisms for the users to digest the uncertainty of the available information [@maceachren1992visualizing, @van1994visualization]. When the needs of the audience are changeable and are also the priority, the map creator can allow interactivity for map users to explore a data set through dynamic interactions. This can allow inspection of the data from many views [@DQBCM]. @TNTEA says user interaction with maps helps them to understand and interpret the spatial distribution of disease, to validate, explain or explore the presented statistics and their relationships to each other.

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features, found in online cancer maps, include tooltips, drop-down menus, data selection, zooming, and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports can be found in various online cancer maps identified by @roberts2019communication.

Animation, in contrast to interactivity, usually involves pre-computing views and showing these in a sequence. @TGA provided an overview of animation in association with the R package `gganimate` [@gganimate], with some focus on maps, explaining transitions between frames in a linear, to provide a fixed narrative to passive users. Animations are used to communicate a message by capturing and directing users' attention. It is most often employed to show changes over time. The controls for basic animation are usually placed outside of the plot space [@TGA], the map image is updated/replaced as the animation progresses.

@CPISACA provide weather maps as a thoroughly developed example of animation of spatial displays to communicate information to the general public. The movement of a weather system will follow a forecasted path, all map users can follow the animated path of the weather system across the geography over a specified period.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(cowplot)
library(png)
library(grid)
```

\section*{Abstract}
<!-- 105 words -->

Cancer atlases communicate cancer statistics over geographic domains. These domains are subdivided by administrative areas such as countries, states or suburbs. When communicating human-related statistics in Australia, the geographic map base highlights the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they are important to consider if they are more densely populated. Alternative map displays can reveal patterns that are not obvious when the same data is shown on a choropleth map. Alternative displays have proven to be successful on online news sites, especially for the communication of election data, where votes are aggregated at state level.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are often delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the count data needs to be merged with population data to appropriately calibrate it to incidence, for example, the rate per 100000 people.  This type of data is collected on a routine basis for public health purposes and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of the disease may be identified and effectively addressed with public health policy and actions.
Daniel Exeter [@SE] recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

@@

# Traditional approaches for cancer map displays {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map [@EI]. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals in communities and organizing communities into representable units [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
The use of choropleths to visualize cancer data [@CPISACA], and the development of these maps for displaying disease data [@DMAHP]. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The data was extracted from the official federal statistics on cancer incidence and deaths [@USInteractive], produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. There is also a spatial outlier -- Utah has a noticeably lower rate than its neighbors. Also, Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection of statistics across regions and the administrative control within regions lends itself to choropleth visualization. 
Atlases date to Haviland's maps in 1875, with UK cancer atlases appearing in 1961 [@HEDP], and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

@@

Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

@@

## Additional considerations
<!-- 479 words -->
Maps cannot tell the entire story. Supplementary graphs and plots are often included to add more depth and information. Additional materials such as tables, graphs, and text explanations, support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, standard deviation. When presenting cancer maps, the intuition derived from maps must be 'validated by rigorous statistical analyses' [@MACM], and the display of the supplementary statistics helps in this regard.

@@

# Limitations of choropleth displays

# Contemporary alternatives to choropleths {#ch:alternatives}

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that white lies may be employed to create useful displays and map creators can draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a valueother than the actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, interpreting spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, Dorling [@ACTUC] considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas [@CTTMB]. 
<!-- @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. DI SAYS I think that focuses on bayesian analysis not on cartograms-->

@@

A contiguous cartogram maintains connectivity of the map regions while areas are altered according to a statistic. This transformation often occurs at the expense of the shape of areas [@CBATCC, @NAC, @TAAM].
Three methods are provided for creating value-by-area cartograms [@ACA]. They implement 'map deformation' to account for the value assigned to each area.
Other methods include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].

@@

He acknowledged the sophistication of contiguous cartograms but critiqued their 'very complex shapes,' he answers this with his implementation of maps created using 'the simplest of all shapes'. 
Circular cartograms use the same simple shapes for every region represented and resize the shapes according to the statistic represented or the population for a base map. This familiar shape may be more effective for understanding the spatial distribution than contiguous cartograms, as the 'nonsense' shapes shown in contiguous cartograms have 'no meaning' after distortions are applied [@NISCC].
To produce a compelling map, a gravity model is applied to avoid overlaps and keep spatial relationships with neighboring areas over many iterations. This implementation can work for up to 'one hundred thousand' areas.

@@

# Limitations of alternative displays {#ch:compare}
<!-- 408 words -->

@@

Some mix of tiling, faceting or even micromaps, which allow some spatial continuity while also zooming into small areas, are good solutions for difficult geographies. Table \ref{tab:methods} summarizes the key criteria for evaluating maps and alternative displays. Suggestions and comments to help map creators best communicate their health data and spatial analyses are provided by Moore and Carpenter [@SAMGIS], Bell et al. [@CPISACA].

@@

Monmonier [@HTLWM] says that interactivity can be used to allow users to explore the map for more information and provides flexibility for the display. The user can toggle between different variables, map views or even multiple realizations of future scenarios [@goodchild1994introduction]. This provides additional mechanisms for the users to digest the uncertainty of the available information [@maceachren1992visualizing, @van1994visualization]. When the needs of the audience are changeable and are also the priority, the map creator can allow interactivity for map users to explore a data set through dynamic interactions. This can allow inspection of the data from many views [@DQBCM]. User interaction with maps helps to understand and interpret the spatial distribution of disease, to validate, explain or explore the presented statistics and their relationships to each other [@TNTEA].

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features, found in online cancer maps, include tooltips, drop-down menus, data selection, zooming, and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports can be found in various online cancer maps [@roberts2019communication].

Animation, in contrast to interactivity, usually involves pre-computing views and showing these in a sequence. An overview of animation in association with the R package `gganimate` [@gganimate], with some focus on maps, explaining transitions between frames in a linear, to provide a fixed narrative to passive users [@TGA]. Animations are used to communicate a message by capturing and directing users' attention. It is most often employed to show changes over time. The controls for basic animation are usually placed outside of the plot space [@TGA], the map image is updated/replaced as the animation progresses.

Weather maps are a thoroughly developed example of animation of spatial displays to communicate information to the general public [@CPISACA]. The movement of a weather system will follow a forecasted path, all map users can follow the animated path of the weather system across the geography over a specified period.
